A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal

### 2020-09-28 Manuscript/2020-MEE-Adams_et_al-Physignal.Rmd
MEE version

---
title: A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal

# Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
author:
  - name: Dean C. Adams
    affiliation: a,2
  - name: Erica K. Baken
    affiliation: a,b
  - name: Michael L. Collyer
    affiliation: b
address:
  - code: a
    address: Department of Ecology, Evolution, and Organismal Biology, Iowa State University, Ames, Iowa, 50010. USA.
  - code: b
    address: Department of Science, Chatham University, Pittsburgh, Pennsylvania, 15232. USA.

corresponding_author:
  - code: 2
    text: "To whom correspondence should be addressed. E-mail: dcadams@iastate.edu"

# For footer text
lead_author_surname: Adams et al.

author_contributions: |
  D.C.A. designed the research; D.C.A., E.K.B., and M.L.C. performed the research and wrote the paper.

## Remove this if not required
conflict_of_interest: |
  The authors declare no conflict of interest.  \hfill\break

  Data deposition: Data for the empirical example may be found on DRYAD: doi:10.5061/dryad.b554m44 and doi:10.5061/dryad.59zw3r23m. R-scripts for simulation tests are found at: https://github.com/deanadams/PhySigCompareZ. Computer code for implementing the two-sample comparison of effect sizes is found in geomorph: https://cran.r-project.org/web/packages/geomorph/index.html \hfill\break

  ^2^To whom correspondence should be addressed. E-mail: dcadams\@iastate.edu

abstract: |
  Macroevolutionary studies frequently characterize the phylogenetic signal in phenotypes, however, analytical tools for comparing the strength of that signal across traits remain largely underdeveloped. Here we evaluate the efficacy of Pagel's $\lambda$ to correctly estimate the strength of phylogenetic signal in phenotypic traits across a range of input values. We find that $\lambda$ behaves as a Bernoulli random variable, where estimates are increasingly skewed at larger and smaller input levels of phylogenetic signal. Further, the precision of $\lambda$ varies with input signal. Another measure, Blomberg's *K*, is more consistent across a range of tree sizes, and exhibits a positive relationship with input levels of phylogenetic signal. However, that relationship is decidedly nonlinear. Thus, neither $\lambda$ nor *K* are suitable as effect sizes for measuring the strength of phylogenetic signal, and comparing that signal across datasets. As an alternative, we propose a standardized effect size based on *K*, ($Z_K$), which measures the strength of phylogenetic signal more reliably than does $\lambda$, and places that signal on a common scale for statistical comparison. We develop tests based on $Z_K$ to provide a mechanism for formally comparing the strength of phylogenetic signal across datasets, in much the same manner as effect sizes may be used to summarize patterns in quantitative meta-analysis. Our approach extends the phylogenetic comparative toolkit to address hypotheses that compare the strength of phylogenetic signal between various phenotypic traits, even when those traits are found in different evolutionary lineages or have different units or scales.

significance: |
  Evolutionary biologists wish to quantify and compare the strength of phylogenetic signal across datasets, but analytical tools for these comparisons are generally lacking. Here we develop a standardized effect size, $Z_K$, which measures the strength of phylogenetic signal on a common statistical scale. We also provide a test statistic, $\hat{Z}_{12}$, for comparing the strength of phylogenetic signal across datasets. We find that two commonly used parameters (Pagel's $\lambda$ and Blomberg's *K*), not converted to effect sizes, are unsuitable for this purpose. Our effect-size procedure enables biologists to quantitatively address hypotheses that compare the strength of phylogenetic signal between various phenotypic traits, even when those traits are found in different evolutionary lineages or have different units or scales.

acknowledgements: |
  We thank E. Glynne and B. Juarez for comments on early drafts of the manuscript. This work was supported in part by NSF grant DBI-1902511 (to D.C.A.) and DBI-1902694 (to M.L.C.). \hfill\break

keywords:
  - phylogenetic signal
  - macroevolution
  - lambda
  - kappa

## must be one of: pnasresearcharticle (usual two-column layout), pnasmathematics (one column layout), or pnasinvited (invited submissions only)
pnas_type: pnasresearcharticle

bibliography: Lambda-Refs.bib
csl: pnas.csl

## change to true to add optional line numbering
lineno: true

output: rticles::pnas_article

---
Investigating macroevolutionary patterns of trait variation requires a phylogenetic perspective, because the shared ancestry of species violates the assumption of independence among trait values that is common for statistical tests [@Felsenstein1985; @HarveyPagel1991]. Accounting for this evolutionary non-independence is the purview of *phylogenetic comparative methods* (PCMs): a suite of analytical tools that condition the data by the phylogenetic relatedness of observations [@Grafen1989; @GarlandIves2000; @Rohlf2001; @MartinsHansen1997; @OMeara_et_al2006;  @Beaulieu_et_al2012; @Adams2014b; @AdamsCollyer2018b]. PCMs are predicated on the notion that phylogenetic signal -- the tendency for closely related species to display similar trait values -- is present in cross-species datasets [@Felsenstein1985; @Pagel1999; @Blomberg_et_al2003]. Indeed, under numerous evolutionary models, phylogenetic signal is expected, as stochastic character change along the hierarchical structure of the tree of life generates trait covariation among taxa [@Felsenstein1985; @Blomberg_et_al2003; @Revell_et_al2008].

Several analytical tools have been developed to quantify phylogenetic signal in phenotypic datasets [@Abouheif1999; @Gittleman1990; @Blomberg_et_al2003; @Adams2014a; @Pagel1999; @Klingenberg2010], and their statistical properties -- namely type I error rates and statistical power -- have been investigated to determine under what conditions phylogenetic signal can be detected [@Munkemuller_et_al2012; @Pavoine2012;  @DinizFilho2012; @Adams2014a; @MolinaVenegas2017; @Revell_et_al2008; @Revell2010; @Boettiger_et_al2012]. One of the most widely used methods for characterizing phylogenetic signal is Pagel's $\lambda$ [@Pagel1999], which transforms the lengths of the internal branches of the phylogeny to improve the fit of data to the phylogeny via maximum likelihood [@Pagel1999; @Freckleton_et_al2002]. When incorporated in PGLS, $\lambda$ serves as a tuning parameter which is optimized via log-likelihood profiling while evaluating the covariation between the dependent and independent variables, given the phylogeny [@Pagel1999; @Freckleton_et_al2002]. To infer whether phylogenetic signal differs from no signal or a Brownian motion model of evolutionary divergence, the observed model fit using $\hat\lambda$ may be statistically compared to that using  $\lambda=0$ or $\lambda=1$ via likelihood ratio tests [@Freckleton_et_al2002; @Cooper2010; @Bose2019] or confidence limits [@Vandelook2019].

Another widely used measure is Blomberg's *K* [@Blomberg_et_al2003], which characterizes phylogenetic signal as the ratio of observed trait variation to the amount of variation expected under Brownian motion. Blomberg's *K* can be treated as a test statistic by employing a permutation test to generate its sampling distribution [@Blomberg_et_al2003; @Adams2014a] for determining whether significant phylogenetic signal is present in data. Both $\lambda$ and *K* seem intuitive to interpret, as a value of $0$ for both corresponds to no phylogenetic signal, while a value of $1$ corresponds to the amount of phylogenetic signal expected under Brownian motion. Thus, it is tempting to regard both $\lambda$ and *K* as descriptive statistics that measure the relative strength of phylogenetic signal, providing an estimate of its magnitude for comparison.

The appeal of Pagel's $\lambda$ and Blomberg's *K* as descriptive statistics is that they provide a basis for interpreting "weak" versus "strong" phylogenetic signal; i.e., small versus large values of $\hat{\lambda}$ or *K*, respectively, in a comparative sense [@DeMeester2019; @Pintanel2019;@Su2019]. Nonetheless, an important question that has yet to be considered is whether such comparisons are analytically appropriate, and whether these statistics are, or can be, converted to effect sizes for comparative analyses across datasets. To be statistics representing phylogenetic signal, they should have reliable distributional properties, which could be revealed with simulation experiments. For instance, as a proportional random variable bounded by $0$ and $1$, we might expect that $\hat{\lambda}$ is a random variable that follows the distribution of a Bernoulli probability parameter [@Forbes2011]; i.e., branch lengths in a tree are scaled proportionally to the probability that data arise from a BM process. Given a known $\lambda$ value used to generate random data on a tree, we would also expect  that the mean of an empirical sampling distribution of $\hat{\lambda}$ would approximately equal $\lambda$; the dispersion of $\hat{\lambda}$ would be largest at intermediate values of $\lambda$, $\hat{\lambda}$ would be predictable over the range of $\lambda$ with respect to tree size; the distribution of $\hat{\lambda}$ would be symmetric at intermediate values of $\lambda$ and more skewed toward values of 0 or 1; and that the distribution of $\hat{\lambda}$ would be more platykurtic at intermediate values of $\lambda$, becoming more leptokurtic toward 0 and 1 [@Forbes2011]. Prior work [@Munkemuller_et_al2012] seems to support some of these conjectures, based superficially on statistical moments for a given tree size (mean, variance, skewness, and kurtosis; see Fig. 2 of ref. [@Munkemuller_et_al2012]). However, because the "strength of Brownian motion" was simulated as a varied weighted-average of data simulated on trees with $\lambda=0$ and $\lambda=1$ and not as prescribed values of $\lambda$ [@Munkemuller_et_al2012], interpretation of these patterns is challenging.

By contrast, for Blomberg's *K*, which is positively unbounded, we might expect that for any $\lambda$ used to generate data, estimates of *K* might be a random variable that follows a normal distribution, with values distributed symmetrically [@Forbes2011]. This attribute seemed less reasonable based on the simulations performed by Münkemüller et al. [-@Munkemuller_et_al2012], which suggested that distributions were positively skewed and that Blomberg's *K* might not behave as a statistic that follows a normal distribution. However, because their simulations used a weighted combination of simulated phylogenetic signal strengths, strong inferences are not possible (and distributional attributes were not the intended result of their simulations). Thus, for both Pagel's $\lambda$ or Blomberg's *K*, evaluation of statistical moments across a range of $\lambda$ used to generate data would be valuable for adjudicating the reliability of these statistics as effect sizes.  Furthermore, the expected values of these statistics appear to vary with tree size [@Munkemuller_et_al2012], making comparisons across studies challenging. Therefore, transformation of these statistics into $Z$-scores would allow evaluation of the efficacy of each statistic to yield effect sizes that could be used for comparisons of the strength of phylogenetic signal across traits and lineages.

Here we use simulation experiments to compare the distributional attributes of $\hat{\lambda}$ and *K*, plus their effect sizes ($Z$-scores), across a range of tree size and phylogenetic signal strength.  We find that estimates of $\hat{\lambda}$ are increasingly skewed at larger and smaller input levels of phylogenetic signal and at smaller tree sizes, vary widely for a given input value of $\lambda$, and that the precision of $\hat{\lambda}$ is not constant across its range. By contrast, estimates of *K* are more consistent across tree sizes, and are normally distributed across the range of input levels of $\lambda$, making *K* a more reliable statistic. We then propose an effect size based on *K*, ($Z_K$), which provides consistent estimates of the strength of phylogenetic signal across tree sizes and signal strength, and facilitates quantitative comparisons of the relative strength of phylogenetic signal across datasets.

# Results

\begin{figure}
\centering
\includegraphics[width=8.25cm]{{fig.1}.png}
\caption{Response of Pagel's \(\lambda\) to increasing strength of
Brownian motion. Gray line signifies the 1:1 line where the input value
matches the estimate \(\hat\lambda\). At each input level, the dark
black line represents the empirically derived expected value (mean) of
\(\hat\lambda\), the red line is the standard deviation of
\(\hat\lambda\), and the blue line is Shapiro Wilks statistic of
\(\hat\lambda\) (\(W=1.0\) signifies normality, \(W< 1.0\) represent
skewed distributions).{}}
\end{figure}

**Lambda ($\lambda$) estimates of phylogenetic signal are inaccurate.** Computer simulations reveal that for $\hat{\lambda}$, the distributional expectations of a Bernoulli variable were mostly upheld. First, the mean value of $\hat{\lambda}$ increases as $\lambda$ increases. Second, the precision in estimating $\lambda$ varies across the range of input values, as the standard deviation of $\hat{\lambda}$ is largest at intermediate values of $\lambda$ and smallest at extreme values (Fig.1 red line). Third, the distributions of $\hat{\lambda}$ tend toward normal distributions at intermediate levels of $\lambda$ but become increasingly skewed at more extreme values of $\lambda$ (Fig. 1 blue line). For small tree sizes, it is also clear that distributions are more platykurtic at intermediate values of $\hat{\lambda}$.  However, the mean value of $\hat{\lambda}$ is negatively-biased (particularly for small tree sizes but also consistently across most of its range; Fig. 1 black line) and standard deviations of $\hat{\lambda}$ are negatively associated with tree size.  For tress of 128 species or less, $\hat{\lambda}$ are quite variable, except for cases when $\lambda$ is near or equal to $1$. Taken together these results reveal that $\hat\lambda$ is a biased statistic that inconsistently estimates  phylogenetic signal, both across tree sizes and across the range of input values. Additional simulations (Supporting Information) reveal that incorporating $\hat\lambda$ in PGLS ANOVA and regression does not adversely affect the statistical properties of PGLS parameter estimation or model evaluation (type I error, power, bias in coefficients). Thus, it is reasonable to incorporate $\hat\lambda$ in PGLS as a parameter for tuning the degree of phylogenetic signal in the dependent variables during the analysis. However, the statistical properties shown in Fig. 1 demonstrate that $\lambda$ is unsuitable as an effect size for measuring the strength of phylogenetic signal in data, and thus $\lambda$ should not be used for comparing phylogenetic signal across datasets.

\begin{figure}
\centering
\includegraphics[width=8.25cm]{{fig.2}.png}
\caption{Response of Blomberg's *K* to increasing strength of Brownian motion.  At each input level, the black line represents the empirically derived expected value (mean) of
*K*, the red line is the standard deviation of
*K*, and the blue line is Shapiro Wilks statistic of
*K* (\(W=1.0\) signifies normality, \(W< 1.0\) represent
skewed distributions).{}}
\end{figure}

**Kappa (*K*) estimates of phylogenetic signal are more reliable.** Simulation results demonstrate that *K* displays better statistical properties. First, as expected, mean values of *K* increase with increasing signal ($\lambda$) irrespective of tree size, albeit non-linearly (Fig. 2 black line). Second, the standard deviation of *K* is consistent across tree sizes (Fig. 2 red line), and while it increases with $\lambda$, it is always less than the mean (low coefficient of variation). This finding is perhaps unsurprising, as *K* is lower-bounded by 0, and is never large for small values of $\lambda$. Importantly, *K* is normally distributed across the range of input $\lambda$; a consistent pattern regardless of tree size (Fig. 2 blue line). This differs from results of [@Munkemuller_et_al2012], where the skewing appears to be due to combining random values generated independently, rather than being a property of *K* itself. Overall, these findings reveal that while *K* is more reliable as an estimate of phylogenetic signal, the non-linear scaling with input signal implies that it should not be considered an effect size that measures the strength of phylogenetic signal on a common scale for comparison across datasets.

**Effect sizes from *K* ($Z_K$) better characterize phylogenetic signal.**  To measure the strength of phylogenetic signal on a common scale, we propose effect sizes (Z-scores) for both $\lambda$ and *K*. Statistically, a standardized effect size may be found as:

@@

where $\theta_{obs}$ is the observed test statistic, $E(\theta)$ is its expected value under the null hypothesis, and $\sigma_\theta$ is its standard error [@Glass1976; @Cohen1988;@Rosenthal1994]. Typically, $\theta_{obs}$ and $\sigma_\theta$ are estimated from the data, while $E(\theta)$ is obtained from the distribution of $\theta$ derived from parametric theory. However, recent advances in resampling theory [@Collyer_et_al2015a;@AdamsCollyer2016; @CollyerAdams2018;@AdamsCollyer2019b] have shown that $E(\theta)$ and $\sigma_\theta$ may also be obtained from an empirical sampling distribution of $\theta$ simulated from permutation procedures.

@@

where $K_{obs}$ is the observed phylogenetic signal, and $\hat\mu_K$ and $\hat\sigma_K$ are the mean and standard deviation of the empirical sampling distribution of *K* obtained via permutation. The empirical sampling distribution of *K* can be first transformed via a Box-Cox transformation to better adhere to the assumption of normality.

@@

where, $\hat{\lambda}$ is the maximized likelihood value of $\lambda$ and $\chi^{2}_{\hat{\lambda}}$ is the likelihood ratio statistic for the value.

@@

\begin{figure}
\centering
\includegraphics[width=8.25cm]{{fig.3}.png}
\caption{Response of effect sizes \(Z_{\lambda}\) and \(Z_K\) to increasing strength of Brownian motion.  Means from simulation runs are shown for comparative ease. Individual values from each simulation run are available in Supporting Informaton. }
\end{figure}

**A test statistic ($\hat{Z}_{12}$) allows meaningful comparisons across datasets.** To statistically compare the strength of phylogenetic signal across datasets we propose a two-sample test statistic ($\hat{Z}_{12}$). Based on statistical theory, a two-sample test statistic may be calculated as:

@@

# Methods

>>>>

---
title: "A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal"

output:
  pdf_document:
    fig_caption: yes
    number_sections: yes
    keep_tex: true
    pandoc_args:

csl: methods-in-ecology-and-evolution.csl
bibliography: Lambda-Refs.bib

header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage{lineno}\linenumbers

---
\begin{center}
\textbf{Dean C. Adams$^{1,*}$, Erica K. Baken$^{1,2}$, and Michael L. Collyer$^2$}
\end{center}

\begin{center}`r format(Sys.time(), '%d %B, %Y')`\end{center}

$^{1}$Department of Ecology, Evolution, and Organismal Biology, Iowa State University, Ames, Iowa, USA.

$^{2}$Department of Science, Chatham University, Pittsburgh, Pennsylvania, USA.

$^{*}$Correspondence: Dean C. Adams <dcadams@iastate.edu> 
\hfill\break

**Keywords**: phylogenetic signal, macroevolution, lambda, kappa \hfill\break

**Short Title**: Effect size for phylogenetic signal \hfill\break

```{r setup, include=FALSE}
  library(knitr)
  knitr::opts_chunk$set(echo = TRUE)
```

```{r, include = FALSE}
library(stringr)
library(tidyverse)

RmdWords <- function(file) {

  # Creates a string of text
  file_string <- file %>%
    readLines() %>%
    paste0(collapse = " ") %>%
    # Remove YAML header
    str_replace_all("^<--- .*?--- ", "") %>%    
    str_replace_all("^--- .*?--- ", "") %>%
    # Remove code
    str_replace_all("```.*?```", "") %>%
    str_replace_all("`.*?`", "") %>%
    # Remove LaTeX
    str_replace_all("[^\\\\]\\$\\$.*?[^\\\\]\\$\\$", "") %>%
    str_replace_all("[^\\\\]\\$.*?[^\\\\]\\$", "") %>%
    # Deletes text between tags
    str_replace_all("TC:ignore.*?TC:endignore", "") %>%
    str_replace_all("[[:punct:]]", " ") %>%
    str_replace_all("  ", "") %>%
    str_replace_all("<", "") %>%
    str_replace_all(">", "")

  # Save several different results
  word_count <- str_count(file_string, "\\S+")
  char_count <- str_replace_all(string = file_string, " ", "") %>% str_count()

   return(list(num_words = word_count, num_char = char_count, word_list = file_string))
}

words <- RmdWords("2020-MEE-Adams_et_al-Physignal.Rmd")
```

**Research Article:**

```{r, echo = FALSE}
count <- c(words = words$num_words, characters = words$num_char)
knitr::kable(t(count))
```

**Author Contributions**:  DCA conceived the original idea for this manuscript, and DCA, EKB, and MLC collaboratively developed the concept and contributed to all portions of this manuscript. All authors approve of the final product and are willingly accountable for any portion of the content.\hfill\break

**Data Archiving**: Empirical data used in this paper are available on DRYAD (associated with original articles). R-scripts for simulation tests are found at: https://github.com/deanadams/PhySigCompareZ. Computer code for implementing the two-sample comparison of effect sizes is found in geomorph (**upon article acceptance**): https://cran.r-project.org/web/packages/geomorph/index.html \hfill\break

**Acknowledgments**: We thank E. Glynne and B. Juarez for comments on early drafts of the manuscript. This work was sponsored in part by National Science Foundation Grants DBI-1902511 (to DCA) and DBI-1902694 (to MLC). The authors have no known conflicts of interest.

\newpage

\setcounter{section}{0}

# Abstract {-}

1. Macroevolutionary studies frequently characterize the phylogenetic signal in phenotypes, however, analytical tools for comparing the strength of that signal across traits remain largely underdeveloped.

2. In this paper, we evaluate the efficacy of Pagel's $\lambda$ to correctly estimate the strength of phylogenetic signal in phenotypic traits across a range of input values. We find that $\lambda$ behaves as a Bernoulli random variable, where estimates are increasingly skewed at larger and smaller input levels of phylogenetic signal. Further, the precision of $\lambda$ varies with input signal.

3. Another measure, Blomberg's *K*, is more consistent across a range of tree sizes, and exhibits a positive relationship with input levels of phylogenetic signal. However, that relationship is decidedly nonlinear. Thus, neither $\lambda$ nor *K* are suitable as effect sizes for measuring the strength of phylogenetic signal, and comparing that signal across datasets.

4. As an alternative, we propose a standardized effect size based on *K*, ($Z_K$), which measures the strength of phylogenetic signal more reliably than does $\lambda$, and places that signal on a common scale for statistical comparison. We develop tests based on $Z_K$ to provide a mechanism for formally comparing the strength of phylogenetic signal across datasets, in much the same manner as effect sizes may be used to summarize patterns in quantitative meta-analysis.

5. Our approach extends the phylogenetic comparative toolkit to address hypotheses that compare the strength of phylogenetic signal between various phenotypic traits, even when those traits are found in different evolutionary lineages or have different units or scales.

\newpage

# Introduction

Investigating macroevolutionary patterns of trait variation requires a phylogenetic perspective, because the shared ancestry of species violates the assumption of independence among trait values that is common for statistical tests [@Felsenstein1985; @HarveyPagel1991]. Accounting for this evolutionary non-independence is the purview of *phylogenetic comparative methods* (PCMs): a suite of analytical tools that condition the data by the phylogenetic relatedness of observations [@Grafen1989; @GarlandIves2000; @Rohlf2001; @MartinsHansen1997; @OMeara_et_al2006;  @Beaulieu_et_al2012; @Adams2014b; @AdamsCollyer2018b]. PCMs are predicated on the notion that phylogenetic signal -- the tendency for closely related species to display similar trait values -- is present in cross-species datasets [@Felsenstein1985; @Pagel1999; @Blomberg_et_al2003]. Indeed, under numerous evolutionary models, phylogenetic signal is expected, as stochastic character change along the hierarchical structure of the tree of life generates trait covariation among taxa [@Felsenstein1985; @Blomberg_et_al2003; @Revell_et_al2008]. \hfill\break

Several analytical tools have been developed to quantify phylogenetic signal in phenotypic datasets [@Abouheif1999; @Gittleman1990; @Blomberg_et_al2003; @Adams2014a; @Pagel1999; @Klingenberg2010], and their statistical properties -- namely type I error rates and statistical power -- have been investigated to determine under what conditions phylogenetic signal can be detected [@Munkemuller_et_al2012; @Pavoine2012;  @DinizFilho2012; @Adams2014a; @MolinaVenegas2017; @Revell_et_al2008; @Revell2010; @Boettiger_et_al2012]. One of the most widely used methods for characterizing phylogenetic signal is Pagel's $\lambda$ [@Pagel1999], which transforms the lengths of the internal branches of the phylogeny to improve the fit of data to the phylogeny via maximum likelihood [@Pagel1999; @Freckleton_et_al2002]. When incorporated in PGLS, $\lambda$ serves as a tuning parameter which is optimized via log-likelihood profiling while evaluating the covariation between the dependent and independent variables, given the phylogeny [@Pagel1999; @Freckleton_et_al2002]. To infer whether phylogenetic signal differs from no signal or a Brownian motion (BM) model of evolutionary divergence, the observed model fit using $\hat\lambda$ may be statistically compared to that using  $\lambda=0$ or $\lambda=1$ via likelihood ratio tests [@Freckleton_et_al2002; @Cooper2010; @Bose2019] or confidence limits [@Vandelook2019]. \hfill\break

Another widely used measure is Blomberg's *K* [@Blomberg_et_al2003], which characterizes phylogenetic signal as the ratio of observed trait variation to the amount of variation expected under Brownian motion. Blomberg's *K* can be treated as a test statistic by employing a permutation test to generate its sampling distribution [@Blomberg_et_al2003; @Adams2014a] for determining whether significant phylogenetic signal is present in data. Both $\lambda$ and *K* seem intuitive to interpret, as a value of $0$ for both corresponds to no phylogenetic signal, while a value of $1$ corresponds to the amount of phylogenetic signal expected under Brownian motion. Thus, it is tempting to regard both $\lambda$ and *K* as descriptive statistics that measure the relative strength of phylogenetic signal, providing an estimate of its magnitude for comparison. \hfill\break

The appeal of Pagel's $\lambda$ and Blomberg's *K* as descriptive statistics is that they provide a basis for interpreting "weak" versus "strong" phylogenetic signal; i.e., small versus large values of $\hat{\lambda}$ or *K*, respectively, in a comparative sense [@DeMeester2019; @Pintanel2019;@Su2019]. Nonetheless, an important question that has yet to be considered is whether such comparisons are analytically appropriate, and whether these statistics are, or can be, converted to effect sizes for comparative analyses across datasets. To be statistics representing phylogenetic signal, they should have reliable distributional properties, which could be revealed with simulation experiments. For instance, as a proportional random variable bounded by $0$ and $1$, we might expect that $\hat{\lambda}$ is a random variable that follows the distribution of a Bernoulli probability parameter [@Forbes2011]; i.e., branch lengths in a tree are scaled proportionally to the probability that data arise from a BM process. Given a known $\lambda$ value used to generate random data on a tree, we would also expect  that the mean of an empirical sampling distribution of $\hat{\lambda}$ would approximately equal $\lambda$; the dispersion of $\hat{\lambda}$ would be largest at intermediate values of $\lambda$, $\hat{\lambda}$ would be predictable over the range of $\lambda$ with respect to tree size; the distribution of $\hat{\lambda}$ would be symmetric at intermediate values of $\lambda$ and more skewed toward values of 0 or 1; and that the distribution of $\hat{\lambda}$ would be more platykurtic at intermediate values of $\lambda$, becoming more leptokurtic toward 0 and 1 [@Forbes2011]. Prior work [@Munkemuller_et_al2012] seems to support some of these conjectures, based superficially on statistical moments for a given tree size (mean, variance, skewness, and kurtosis; see Fig. 2 of ref. [@Munkemuller_et_al2012]). However, because the "strength of Brownian motion" was simulated as a varied weighted-average of data simulated on trees with $\lambda=0$ and $\lambda=1$ and not as prescribed values of $\lambda$ [@Munkemuller_et_al2012], interpretation of these patterns is challenging. \hfill\break

By contrast, for Blomberg's *K*, which is positively unbounded, we might expect that for any $\lambda$ used to generate data, estimates of *K* might be a random variable that follows a normal distribution, with values distributed symmetrically [@Forbes2011]. This attribute seemed less reasonable based on the simulations performed by Münkemüller et al. [-@Munkemuller_et_al2012], which suggested that distributions were positively skewed and that Blomberg's *K* might not behave as a statistic that follows a normal distribution. However, because their simulations used a weighted combination of simulated phylogenetic signal strengths, strong inferences are not possible (and distributional attributes were not the intended result of their simulations). Thus, for both Pagel's $\lambda$ or Blomberg's *K*, evaluation of statistical moments across a range of $\lambda$ used to generate data would be valuable for adjudicating the reliability of these statistics as effect sizes.  Furthermore, the expected values of these statistics appear to vary with tree size [@Munkemuller_et_al2012], making comparisons across studies challenging. Therefore, transformation of these statistics into $Z$-scores would allow evaluation of the efficacy of each statistic to yield effect sizes that could be used for comparisons of the strength of phylogenetic signal across traits and lineages. \hfill\break

Here we use simulation experiments to compare the distributional attributes of $\hat{\lambda}$ and *K*, plus their effect sizes ($Z$-scores), across a range of tree size and phylogenetic signal strength.  We find that estimates of $\hat{\lambda}$ are increasingly skewed at larger and smaller input levels of phylogenetic signal and at smaller tree sizes, vary widely for a given input value of $\lambda$, and that the precision of $\hat{\lambda}$ is not constant across its range. By contrast, estimates of *K* are more consistent across tree sizes, and are normally distributed across the range of input levels of $\lambda$, making *K* a more reliable statistic. However, the relationship between *K* and input levels of phylogenetic signal is decidedly nonlinear. Thus, neither $\lambda$ nor *K* are suitable as effect sizes for measuring the strength of phylogenetic signal. As an alternative, we propose an effect size based on *K*, ($Z_K$), which provides consistent estimates of the strength of phylogenetic signal across tree sizes and signal strength. Further, because $Z_K$ places that phylogenetic signal on a common scale, it facilitates statistical comparisons of the relative strength of phylogenetic signal across datasets. We propose a two-sample statistic ($\hat{Z}_{12}$) to accomplish this task, and show that it displays appropriate levels of type I error and model misspecification. An empirical example is then provided to illustrate its use.

# Simulation methods and results

Methods for characterizing phylogenetic signal were evaluated by computer simulation. Briefly, simulations were conducted by generating pure-birth phylogenies at each of six different tree sizes ($n=2^5, 2^6, \cdots, 2^{10}$), and with differing levels of phylogenetic signal ($\lambda=0.0, 0.5, \cdots, 1.0$). We generated 50 random trees for each intersection of tree size and $\lambda$. For each $\lambda$ within each tree size, continuous traits were then simulated on each phylogeny under a BM model of evolution. For each set of 50 trees we measured the mean values of $\hat{\lambda}$ and *K*, their standard deviation, and calculated the Shapiro-Wilk $W$ statistic as a departure from normality (symmetry). For the latter, a value of $1.0$ indicates normally distributed values, while departures from $1.0$ indicate skewness. Simulations were then repeated for both balanced and pectinate trees, which yielded qualitatively similar results (see Supporting Information). Trees containing polytomies, and an evaluation of $\hat{\lambda}$ from models of linear regression and phylogenetic ANOVA, were also investigated, and results were qualitatively similar to those reported above (see Supporting Information). \hfill\break

## Lambda ($\lambda$) estimates of phylogenetic signal are inaccurate

Computer simulations reveal that for $\hat{\lambda}$, the distributional expectations of a Bernoulli variable were mostly upheld. First, the mean value of $\hat{\lambda}$ increases as $\lambda$ increases. Second, the precision in estimating $\lambda$ varies across the range of input values, as the standard deviation of $\hat{\lambda}$ is largest at intermediate values of $\lambda$ and smallest at extreme values (Fig.1 red line). Third, the distributions of $\hat{\lambda}$ tend toward normal distributions at intermediate levels of $\lambda$ but become increasingly skewed at more extreme values of $\lambda$ (Fig. 1 blue line). For small tree sizes, it is also clear that distributions are more platykurtic at intermediate values of $\hat{\lambda}$.  However, the mean value of $\hat{\lambda}$ is negatively-biased (particularly for small tree sizes but also consistently across most of its range; Fig. 1 black line) and standard deviations of $\hat{\lambda}$ are negatively associated with tree size.  For tress of 128 species or less, $\hat{\lambda}$ are quite variable, except for cases when $\lambda$ is near or equal to $1$. Taken together these results reveal that $\hat\lambda$ is a biased statistic that inconsistently estimates  phylogenetic signal, both across tree sizes and across the range of input values. Additional simulations (Supporting Information) reveal that incorporating $\hat\lambda$ in PGLS ANOVA and regression does not adversely affect the statistical properties of PGLS parameter estimation or model evaluation (type I error, power, bias in coefficients). Thus, it is reasonable to incorporate $\hat\lambda$ in PGLS as a parameter for tuning the degree of phylogenetic signal in the dependent variables during the analysis. However, the statistical properties shown in Fig. 1 demonstrate that $\lambda$ is unsuitable as an effect size for measuring the strength of phylogenetic signal in data, and thus $\lambda$ should not be used for comparing phylogenetic signal across datasets.

## Kappa (*K*) estimates of phylogenetic signal are more reliable

Simulation results demonstrate that *K* displays better statistical properties. First, as expected, mean values of *K* increase with increasing signal ($\lambda$) irrespective of tree size, albeit non-linearly (Fig. 2 black line). Second, the standard deviation of *K* is consistent across tree sizes (Fig. 2 red line), and while it increases with $\lambda$, it is always less than the mean (low coefficient of variation). This finding is perhaps unsurprising, as *K* is lower-bounded by 0, and is never large for small values of $\lambda$. Importantly, *K* is normally distributed across the range of input $\lambda$; a consistent pattern regardless of tree size (Fig. 2 blue line). This differs from results of [@Munkemuller_et_al2012], where the skewing appears to be due to combining random values generated independently, rather than being a property of *K* itself. Overall, these findings reveal that while *K* is more reliable as an estimate of phylogenetic signal, the non-linear scaling with input signal implies that it should not be considered an effect size that measures the strength of phylogenetic signal on a common scale for comparison across datasets.

## Effect sizes from *K* ($Z_K$) better characterize phylogenetic signal

To measure the strength of phylogenetic signal on a common scale, we propose effect sizes (Z-scores) for both $\lambda$ and *K*. Statistically, a standardized effect size may be found as:

@@

where $\theta_{obs}$ is the observed test statistic, $E(\theta)$ is its expected value under the null hypothesis, and $\sigma_\theta$ is its standard error [@Glass1976; @Cohen1988;@Rosenthal1994]. Typically, $\theta_{obs}$ and $\sigma_\theta$ are estimated from the data, while $E(\theta)$ is obtained from the distribution of $\theta$ derived from parametric theory. However, recent advances in resampling theory [@Collyer_et_al2015a;@AdamsCollyer2016; @CollyerAdams2018;@AdamsCollyer2019b] have shown that $E(\theta)$ and $\sigma_\theta$ may also be obtained from an empirical sampling distribution of $\theta$ simulated from permutation procedures. \hfill\break

@@

where $K_{obs}$ is the observed phylogenetic signal, and $\hat\mu_K$ and $\hat\sigma_K$ are the mean and standard deviation of the empirical sampling distribution of *K* obtained via permutation. The empirical sampling distribution of *K* can be first transformed via a Box-Cox transformation to better adhere to the assumption of normality. \hfill\break

@@

where, $\hat{\lambda}$ is the maximized likelihood value of $\lambda$ and $\chi^{2}_{\hat{\lambda}}$ is the likelihood ratio statistic for the value. \hfill\break

@@

## A test statistic ($\hat{Z}_{12}$) allows meaningful comparisons across datasets

To statistically compare the strength of phylogenetic signal across datasets we propose a two-sample test statistic ($\hat{Z}_{12}$). Based on statistical theory, a two-sample test statistic may be calculated as:

@@

\newpage

# References {-}

 \setlength{\parindent}{-0.25in}
 \setlength{\leftskip}{0.25in}
 \setlength{\parskip}{8pt}
 \noindent

<div id="refs"></div>

\newpage

# Figures

Figure 1. Response of Pagel's $\lambda$ to increasing strength of Brownian motion. Gray line signifies the 1:1 line where the input value matches the estimate $\hat\lambda$. At each input level, the dark black line represents the empirically derived expected value (mean) of $\hat\lambda$, the red line is the standard deviation of $\hat\lambda$, and the blue line is Shapiro Wilks statistic of $\hat\lambda$ ($W=1.0$ signifies normality, $W< 1.0$ represent skewed distributions).  \hfill\break

Figure 2. Response of Blomberg's $K$ to increasing strength of Brownian motion.  At each input level, the black line represents the empirically derived expected value (mean) of $K$, the red line is the standard deviation of *K*, and the blue line is Shapiro Wilks statistic of $K$ ($W=1.0$ signifies normality, $W< 1.0$ represent skewed distributions). \hfill\break
